---
title: "Boxplot Tutorial"
author: "Varsha Venkatesh"
date: "2026-07-14"
output: html_document
---
## Getting started

To create a boxplot in R, you will need a few packages and a dataset with both a categorical and continuous variable.

Let's say we want to make a boxplot comparing term GPAs by student major during the first semester of college.

```{r setup, include = TRUE}
# for data manipulation
library(tidyr)

# for boxplot
library(ggplot2)
library(jtools) # to put diagrams in apa format
```

```{r data, include = TRUE}
# set working directory to iwell_tutorials/Tutorials
dat <- read.csv("../Data/iwell_simulated_data_2026-07-14.csv")
```

## Data manipulation

Our dataset "dat" contains missing data, meaning ggplot2 will automatically make a boxplot for anyone whose major is labeled as NA. We don't want this, so we will remove anyone whose major in semester 1 is NA.

```{r data manipulation, include = TRUE}
# making a new dataset and removing NAs
boxplotdat <- dat %>% 
  filter(!is.na(major_sem_1)) # only keeps participants who do not have NAs in the "major_sem_1" column
```

## Making diagram

Now that we've loaded in our packages and dataset, we can make our boxplot!

```{r data manipulation, include = TRUE}
boxplot <- ggplot(boxplotdat, aes(major_sem_1, gpa_sem_1, fill = major_sem_1)) +
  geom_boxplot() + 
  theme_apa() +
  theme(legend.position = "bottom") +
  ggtitle("First Semester GPA by College Major") +
  xlab("Major") +
  ylab("Term GPA") +
  scale_fill_manual(values = c("#f94144", "#f9c74f", "#f8961e", "#577590", "#90be6d"))

# exporting plot
ggsave(
  filename = file.path("../Figures", "boxplot_example.png"), 
  plot = boxplot,
  width = 11,
  height = 8.5,
  units = "in",    
  dpi = 300
)
```